---
title: "GEOG 616 Lab 3"
author: "Griffin Shelor"
format: html
---

### Question 1
Using the sf package, A) Create a sf points object by adding 100 random x and y values between 0 and 10 with a uniform distribution (runif) to the beaver2 dataset (which is acquired by loading data(beavers) from the "datasets" package). B) Show the structure of your dataset.

```{r 1A & 1B}
setwd("/Users/griffinshelor/Documents/Schoolwork/UNR/GEOG616/Labs/Lab3")
library(datasets)
library(sf)
library(tidyverse)
set.seed(802)
data(beavers)
beaver_2 <- beaver2 |>
  mutate(x_geo = runif(100,0,10),
         y_geo = runif(100,0,10))
beaver_2 <- st_as_sf(beaver_2, coords = c('x_geo', 'y_geo'))
str(beaver_2)
```

### Question 2

With beaver1 from the same dataset that you used in Question 1, A) create a sf points layer by adding random uniform x values between 239652 and 765537 and random uniform y values between 3875802 and 4653617 to the dataset. B) Summarize your object Use your favorite number as your random seed.

```{r 2A & 2B}
set.seed(802)
beaver_1 <- beaver1 |>
  mutate(x_geo = runif(114, 239652, 765537),
         y_geo = runif(114, 3875802, 4653617))
beaver_1 <- st_as_sf(beaver_1, coords = c('x_geo', 'y_geo'))
str(beaver_1)
summary(beaver_1)
```

### Question 3

Find the PROJ 4 projection string for UTM NAD83 Zone 11N (without a specific use area indicator (e.g. HARN) that applies to "North America" at http://spatialreference.org A)Assign this projection to the points object you created in question 2, and B) display the structure of your spatial points object below.

```{r 3A & 3B}
st_crs(beaver_1) <- 26911
str(beaver_1)
```

### Question 4

A) Load the "Nevada Counties" shapefile provided. B) List the projection of this dataset and C) plot the SPDF
```{r 4A, 4B, 4C}
NV_counties <- st_read("NV_County_Boundaries_shp/NV_County_Boundaries.shp")
st_crs(NV_counties)
plot(NV_counties)
```

### Question 5

A) Reproject the Counties shapefile that you loaded in question 4 to North American Datum 1983 lat long using epsg:4269. B) Reproject the points from Question 3 to the same projection. C) Plot the points on top of the Counties excluding points that do not occur in Nevada. D) What are the range of latitudes and longitudes given the new county projection?

```{r 5A & 5B}
NV_counties_4269 <- st_transform(NV_counties, 4269)
beaver_1_4269 <- st_transform(beaver_1, 4269)
```

```{r 5C & 5D}
plot(st_geometry(NV_counties_4269))
plot(st_geometry(beaver_1_4269[NV_counties_4269,]), add = TRUE)
st_geometry(NV_counties_4269)
## The minimum latitude is 35.06266 and the max latitude is 41.94892. The minimum longitude is -119.9168 and the max longitude is -114.052.
```

### Question 6

Repeat the workflow (point creation, re-projection to ll, subsetting by polygon, and plotting) using the terra vector features
```{r}
library(terra)
set.seed(802)
## beaver 2 first
beaver_2terra <- beaver2 |>
  mutate(x_geo = runif(100,0,10),
         y_geo = runif(100,0,10))
beaver_2terra <- vect(beaver_2terra, geom = c('x_geo', 'y_geo'), crs = 'EPSG:26911')
str(beaver_2terra)
## now beaver 1
beaver_1terra <- beaver1 |>
  mutate(x_geo = runif(114, 239652, 765537),
         y_geo = runif(114, 3875802, 4653617))
beaver_1terra <- vect(beaver_1terra, geom = c('x_geo', 'y_geo'), crs = 'EPSG:26911')
str(beaver_1terra)
summary(beaver_1terra)

## reading in NV counties
NV_counties_terra <- vect("NV_County_Boundaries_shp/NV_County_Boundaries.shp", crs = 'EPSG:26911')

## changing the projections
project(NV_counties_terra,'EPSG:4269')
project(beaver_1terra, 'EPSG:4269')

terra::plot(NV_counties_terra, box = F, axes = F)
points(beaver_1terra[NV_counties_terra,])

```


### Question 7

Grad students = YES, Undergraduates = Extra Credit

Read the articles by Brewer 2006 and Yildirim and Kaya 2008 and https://source.opennews.org/articles/choosing-right-map-projection/
What are the implications of choosing an inappropriate projection for analyzing your research project?

Answer: The implications of choosing an inappropriate projection for analyzing my research project will vary depending on what my research project is. If I am mapping tree canopy change in a city or county in Ohio, I would likely want to use a UTM projection, as it is best suited to smaller areas. Even if it is not designated as an "equal area projection", its specificity to my area of study would mean that errors in area calculations would still be limited, thus making the tree canopy change data more precise when I hand it over to policymakers. Choosing an inappropriate projection in this instance would mean that the data I hand over would have errors and thus would not result in the desired policy outcomes of improving a municipality's green infrastructure. However, if I am working on land cover mapping over the whole United States, an equal area projection would be much more appropriate, as a UTM projection would become more distorted the further away my data got from whichever UTM zone I had chosen. If I am creating a navigation chart or a weather map, a conformal projection would be best, as it would protect from distortion of angles and preserve direction. This is important because a less appropriate direction would potentially result in my ship making a wrong turn or accidentally running onto shore as the map being used to navigate would not properly represent which direction the ship was heading in. In the case of the weather map, it would be important for a typical resident of an area to confidently understand whether a storm or other dangerous weather is present in their region and which direction it is heading in. An inappropriate projection might result in residents making unsafe choices that put themselves and others in harm's way due to a miscommunication in where dangerous weather was headed. 
