---
title: "Lab 10 Space and Time"
author: "Griffin Shelor"
date: "11/9/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(zoo)
library(spacetime)
library(xts)
library(tidyverse)
library(gstat)
library(sf)
library(RColorBrewer)
library(maptools)
library(maps)
```

The file NOAA_df_1990.csv contains data on the max temperature taken at weather stations over a 4 year period

Question 1. a) Read in the data. b) Create a new column in the dataset combining the time elements as a POSIXct data class. c) What is the Min and Max date/time that were recorded in this data set?
```{r}
setwd("/Users/griffinshelor/Documents/Schoolwork/UNR/GEOG616/Labs/Lab10")
NOAA_df <- read.csv("NOAA_df_1990.csv")
NOAA_df$date_posix <- as.POSIXct(NOAA_df$date, format = "%Y-%m-%d")
head(NOAA_df)
min(NOAA_df$date_posix)
max(NOAA_df$date_posix)

## The min date is 1990-01-01 and the max date is 1993-12-31.
```

Question 2. 1) Create a spacetime dataframe from your object for the month of July in 1992 and 2) Show the structure and 3) plot the time series and 4) spatial slot.

```{r}
NOAA_df_july <- NOAA_df |>
  filter(month == 7 & year == 1992)
NOAA_stdf_july <- stConstruct(NOAA_df_july, space = c(9, 10), time = 'date_posix')
str(NOAA_stdf_july)
plot(NOAA_stdf_july@sp)
plot(NOAA_stdf_july)
```

Question 3. The data contain several measurements (z) and the type of measurement is given in the proc column. 1) Subset your data to only the Tmax data. 2)  Plot your data in 4 time intervals, with the Tmax variable shown in 5 increments. What are the min and max Tmax values?


```{r}
NOAA_df_julytmax <- NOAA_df_july |>
  filter(proc == "Tmax")
NOAA_stdf_julytmax <- stConstruct(NOAA_df_julytmax, space = c(9, 10), time = 'date_posix')
stplot(NOAA_stdf_julytmax, number = 4, cuts = 5)
summary(NOAA_stdf_julytmax)

## The min and max Tmax values are 57 and 102.
```

Question 4. What is the average of maximum temperature for the last week in July?

```{r}
NOAA_df_julytmax_lastweek <- NOAA_df_julytmax |>
  filter(day > 23)
mean(NOAA_df_julytmax_lastweek$z)

## The average of max temps for the last week in July is 83.53853.
```

Question 5 - a) Construct a time series object of the dew point temperature data (TDP) for the entire month of July of 1992. b) use the Auto and Cross Covariance function (acf) to estimate the correlation associated with time lag in days in which the data are autocorrelated.
```{r}
NOAA_df_julyTDP <- NOAA_df_july |>
  filter(proc == "TDP") |>
  select(date_posix, z)
julyTDP_ts <- ts(NOAA_df_julyTDP)
julyTDP_ts_acf <- acf(julyTDP_ts, plot = F)

plot(julyTDP_ts_acf$acf[,2,2], xlab = 'Lag', ylab = 'ACF', main = 'TDP Time Autocorrelation', type='h')
```


## Question 6 Graduate students - XC for undergrads 
Using the time series object of the dew point temperature data (TDP) for the entire month of July of 1992 that you created in Question 5. A) Create a space-time variogram, B) fit the nugget range and sill for the space and time components. C) create a prediction grid of a 20x20 lat long size, and  dates using the code here

" as.POSIXct(as.Date("1993-07-01") + seq(3, 28, length = 6), tz='PDT') "
 
 Krige and plot your result
 
 
```{r}
set.seed(802)
NOAA_df_julyTDP <- NOAA_df_july |>
  filter(proc == "TDP")
julyTDP_stdf <- stConstruct(NOAA_df_julyTDP, space = c(9, 10), time = 12)
class(julyTDP_stdf)
julyTDP_vg <- variogram(object = z ~ 1,
                        data = julyTDP_stdf,
                        tunit = "days",
                        tlags = 0.1:20)

plot(julyTDP_vg)
```
```{r}
sepVgm <- vgmST(stModel = "separable",
space = vgm(30, "Exp", 400, nugget = 0.1),
time = vgm(30, "Exp", 5, nugget = 0.1),
sill = 45)
sepVgm <- fit.StVariogram(julyTDP_vg, sepVgm)
plot(julyTDP_vg, sepVgm, main = "Semi-variance")

bbox(julyTDP_stdf)

spat_pred_grid <- expand.grid(
lon = seq(-100, -80, length = 20),
lat = seq(32, 46, length = 20)) |>
SpatialPoints(proj4string = CRS(proj4string(julyTDP_stdf)))
class(spat_pred_grid)
plot(spat_pred_grid, pch=19, cex = 0.5)

gridded(spat_pred_grid) <- TRUE
plot(spat_pred_grid)

temp_pred_grid <- as.POSIXct(as.Date("1993-07-01") + seq(3, 28, length = 6), tz="PDT")
temp_pred_grid

DE_pred <- STF(sp = spat_pred_grid,# spatial part
time = temp_pred_grid)# temporal part
class(DE_pred)
summary(DE_pred)
plot(DE_pred)

plot(DE_pred@sp)

# julyTDP_stdf_v2 <- as(julyTDP_stdf[, -14], "STIDF")# convert to STIDF - omitting one problematic column
# julyTDP_stdf_v2 <- subset(julyTDP_stdf_v2, !is.na(julyTDP_stdf_v2$z))

pred_kriged <- krigeST(z ~ 1,
data = julyTDP_stdf, 
newdata = DE_pred,
modelList = sepVgm,
computeVar = TRUE)

color_pal <- rev(colorRampPalette(brewer.pal(11, "Spectral"))(16))
stplot(pred_kriged,
main = "Predictions (degrees Fahrenheit)",
layout = c(3, 2),
col.regions = color_pal)
```


